---
title: "MCR_2021_Patientcohort_data"
author: "Lauren Jillson / James Costello"
date: "3/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Contents:

- [Prepare the data](#inputdata)
- [Annotate Gleason Scores](#annotategleason)
- [TCGA](#tcga)
- [Table 1: Patient Characteristics](#table1)
- [Table 2: CHD1 & MAP3K7 HScores across patients](#table2)
- [Table 3: ERG status across patients](#table3)
- [Compare Hscores in BCR patients](#hscorerelapse)
- [Compare Hscores for ICC vs. NCR patients](#nc4)
- [Table 4: Univariate Logistic Regression](#univariate)
- [Table 5: Multivariate Logistic Regression](#multivariate)
- [Predictor using HScores](#predictor)

&nbsp;
&nbsp;

You will need the following packages:\
**caret**\
**gplots2**\
**gridExtra**\
**ggpubr**\
**cgdsr**


<a name="inputdata"/>

### Read in, clean, and prepare data for analysis

There are 4 files with associated variables, including the clinical features from the Stanford cohort (), H-score measurements from the scoring of MAP3K7, CHD1, and ERG, and the cribriform status. Read in the files, remove patients with empty values, and order the data for further analysis below.
```{r, message=F, echo=F}


# read in the Hscore data.  Note there are may be multiple readings per patient that have multiple samples. The samples used in the analysis are labled with a 1.
hscores <- read.table("HScores.txt",sep="\t", header=T)
hscores.1 <- hscores[hscores$byperson.1.use.for.by.person.analysis ==1,] # samples to include in the analysis
hscores.0 <- hscores[hscores$byperson.1.use.for.by.person.analysis ==0,] # remaining samples

# remove any samples with missing Hscores for ERG, CHD1 or MAP3K7
hscores.1 <- hscores.1[!is.na(hscores.1$HScoreERG),] 
hscores.1 <- hscores.1[!is.na(hscores.1$HScoreCHD1),]
hscores.1 <- hscores.1[!is.na(hscores.1$HScoreMAP3K7),]
row.names(hscores.1) <- hscores.1$ID

# read in Cribriform information
crib <- read.table("Cribriform_summary.txt", sep="\t", header=T, row.names=1)
pats <- intersect(row.names(hscores.1),row.names(crib))
crib <- crib[pats,]
crib$Cribriform <- ifelse(grepl("Y",crib$Cribriform),1,0)
hscores.1 <- hscores.1[pats,]
hscores.1 <- cbind(hscores.1,crib)

# read in biochemical relapse information
bcr <- read.table("BCR_cases_clinical_information.txt", sep="\t", header=T, row.names=1)
bcr.pats <- (intersect(row.names(bcr), row.names(hscores.1)))
bcr <- bcr[bcr.pats,]
hscore.1.bcr <- hscores.1[bcr.pats,]

nobcr <- read.table("NoBCR_controls_clinical_information.txt", sep="\t", header=T, row.names=1)
nobcr.pats <- intersect(row.names(nobcr), row.names(hscores.1))
nobcr <- nobcr[nobcr.pats,]
hscore.1.nobcr <- hscores.1[nobcr.pats,]

```

<a name="annotategleason"/>

### Annotate Gleason scores

The Gleason scores are in the old form, which are percentages of sample in Gleason grade 3,4, or 5.  This section transforms to the more updated X+X Gleason scores.
```{r, message=FALSE, echo=FALSE}
#translate to gleason
GleasonScore <- function(g3, g4, g5) {
  if (g3 > g4 & g3 > g5) {
    if (g3 == 100) {
      return("3+3")
    } else if (g4 > g5) {
      return("3+4")
    } else {
      return("3+5")
    }
  } else if (g4 > g3 & g4 > g5) {
    if (g4 == 100) {
      return("4+4")
    } else if (g3 > g5) {
      return("4+3")
    } else {
      return ("4+5")
    }
  } else {
    if (g5 == 100){
      return("5+5")
    } else if(g4 > g3) {
      return("5+4")
    } else {
      return("5+3")
    }
  }
}



#Translate Gleason scores and then compile to 8+, 4+3, and 3+3 and 3+4
a <- array()
for (i in 1:nrow(bcr)) {
  a[i] <- GleasonScore(bcr$Grade.3[i], bcr$Grade.4[i], bcr$Grade.5[i])
}
bcr$Gleason <- a
bcr$Gleason <- gsub("4\\+4", "8\\+",  bcr$Gleason)
bcr$Gleason <- gsub("4\\+5", "8\\+",  bcr$Gleason)
bcr$Gleason <- gsub("5\\+4", "8\\+",  bcr$Gleason)
bcr$Gleason <- gsub("5\\+3", "8\\+",  bcr$Gleason)
bcr$Gleason <- gsub("3\\+3", "temp",  bcr$Gleason)
bcr$Gleason <- gsub("3\\+4", "temp",  bcr$Gleason)
bcr$Gleason <- gsub("temp", "3\\+3 and 3\\+4",  bcr$Gleason)
#print("Summary of BCR")
#table(bcr$Gleason)

a <- array()
for (i in 1:nrow(nobcr)) {
  a[i] <- GleasonScore(nobcr$Grade.3[i], nobcr$Grade.4[i], nobcr$Grade.5[i])
}
nobcr$Gleason <- a
nobcr$Gleason <- gsub("4\\+4", "8\\+",  nobcr$Gleason)
nobcr$Gleason <- gsub("4\\+5", "8\\+",  nobcr$Gleason)
nobcr$Gleason <- gsub("5\\+4", "8\\+",  nobcr$Gleason)
nobcr$Gleason <- gsub("5\\+3", "8\\+",  nobcr$Gleason)
nobcr$Gleason <- gsub("3\\+3", "temp",  nobcr$Gleason)
nobcr$Gleason <- gsub("3\\+4", "temp",  nobcr$Gleason)
nobcr$Gleason <- gsub("temp", "3\\+3 and 3\\+4",  nobcr$Gleason)
#print("Summary of No BCR")
#table(nobcr$Gleason)
```

<a name="tcga"/>

###  TCGA: RNAseq analysis of CHD1, MAP3K7 and ERG status.

determine the relationsihps of CHD1 and MAP3K7 based on ERG status.

```{r, message=F, echo=F}
library(ggplot2)
library(ggpubr)
library(cgdsr)

mycgds = CGDS("https://www.cbioportal.org/")

# TCGA
i <- match("prad_tcga_pub", getCancerStudies(mycgds)[,1])
mycancerstudy = getCancerStudies(mycgds)[i,1]
j <- match("prad_tcga_pub_rna_seq_v2_mrna", getCaseLists(mycgds,mycancerstudy)[,1])
mycaselist = getCaseLists(mycgds,mycancerstudy)[j,1]
k <- match("prad_tcga_pub_rna_seq_v2_mrna_median_Zscores", getGeneticProfiles(mycgds,mycancerstudy)[,1])
mygeneticprofile = getGeneticProfiles(mycgds,mycancerstudy)[k,1]
tcga.exprs <- getProfileData(mycgds,c('ERG','MAP3K7', 'CHD1'),mygeneticprofile,mycaselist)
myclinicaldata = getClinicalData(mycgds,mycaselist)

i <- match("prad_tcga", getCancerStudies(mycgds)[,1])
mycancerstudy = getCancerStudies(mycgds)[i,1]
j <- match("prad_tcga_rna_seq_v2_mrna", getCaseLists(mycgds,mycancerstudy)[,1])
mycaselist = getCaseLists(mycgds,mycancerstudy)[j,1]
myclinicaldata_v2 = getClinicalData(mycgds,mycaselist)

tmp<- intersect(row.names(myclinicaldata), row.names(tcga.exprs))
pats <- intersect(row.names(myclinicaldata_v2),tmp)
myclinicaldata <- myclinicaldata[pats,]
myclinicaldata_v2 <- myclinicaldata_v2[pats,]
tcga.exprs <- tcga.exprs[pats,]
tcga.exprs <- cbind(tcga.exprs, subtype=myclinicaldata$SUBTYPE, BCR =myclinicaldata_v2$BIOCHEMICAL_RECURRENCE_INDICATOR)
tcga.exprs <- tcga.exprs[tcga.exprs$BCR != "",]
tcga.exprs <- tcga.exprs[tcga.exprs$subtype == "1-ERG" | tcga.exprs$subtype == "8-other" | tcga.exprs$subtype == "7-IDH1" | tcga.exprs$subtype == "6-FOXA1" | tcga.exprs$subtype == "5-SPOP",]
tcga.exprs$subtype <- gsub("1-ERG", "ERG+",  tcga.exprs$subtype)
tcga.exprs$subtype <- gsub("8-other", "ERG-",  tcga.exprs$subtype)
tcga.exprs$subtype <- gsub("7-IDH1", "ERG-",  tcga.exprs$subtype)
tcga.exprs$subtype <- gsub("6-FOXA1", "ERG-",  tcga.exprs$subtype)
tcga.exprs$subtype <- gsub("5-SPOP", "ERG-",  tcga.exprs$subtype)
tcga.exprs$BCR <- gsub("YES", "BCR",  tcga.exprs$BCR)
tcga.exprs$BCR <- gsub("NO", "NOBCR",  tcga.exprs$BCR)

#MAP3K7
p <- ggplot(tcga.exprs, aes(x=BCR, y=MAP3K7)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(height=0, width=.3, cex=2) +
  theme_bw() + 
  ggtitle("MAP3K7 RNAseq (TCGA)") + 
  xlab("Biochemical Relapse") +
  stat_compare_means(method="wilcox.test")+
  ylab("MAP3K7 Expression (z-score)") +
  ylim(c(-4,4))
p

pval.ergpos <- round(wilcox.test(tcga.exprs$MAP3K7[tcga.exprs$BCR == "BCR" & tcga.exprs$subtype == "ERG+"], tcga.exprs$MAP3K7[tcga.exprs$BCR == "NOBCR" & tcga.exprs$subtype == "ERG+"])$p.val,5)
pval.ergneg <- round(wilcox.test(tcga.exprs$MAP3K7[tcga.exprs$BCR == "BCR" & tcga.exprs$subtype == "ERG-"], tcga.exprs$MAP3K7[tcga.exprs$BCR == "NOBCR" & tcga.exprs$subtype == "ERG-"])$p.val,5)
p <- ggplot(tcga.exprs, aes(x=BCR, y=MAP3K7, fill=subtype)) + 
  geom_point(position = position_jitterdodge(.2)) + 
  geom_boxplot(outlier.shape = NA, alpha=0.4) + 
  ggtitle("MAP3K7 RNAseq (TCGA)") + 
  ylab("MAP3K7 Expression (z-score)") +
  xlab("Biochemical Relapse") +
  theme_bw() +
  ylim(c(-4,4)) +
  annotate("text", x = 1.5, y = 3.75, label = paste("Wilcoxon (ERG pos), p=",pval.ergpos, "; (ERG neg) p=",pval.ergneg, sep=""))
p

# CHD1
p <- ggplot(tcga.exprs, aes(x=BCR, y=CHD1)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(height=0, width=.3, cex=2) +
  theme_bw() + 
  ggtitle("CHD1 RNAseq (TCGA)") + 
  xlab("Biochemical Relapse") +
  stat_compare_means(method="wilcox.test")+
  ylab("CHD1 Expression (z-score)") +
  ylim(c(-4,4))
p

pval.ergpos <- round(wilcox.test(tcga.exprs$CHD1[tcga.exprs$BCR == "BCR" & tcga.exprs$subtype == "ERG+"], tcga.exprs$CHD1[tcga.exprs$BCR == "NOBCR" & tcga.exprs$subtype == "ERG+"])$p.val,5)
pval.ergneg <- round(wilcox.test(tcga.exprs$CHD1[tcga.exprs$BCR == "BCR" & tcga.exprs$subtype == "ERG-"], tcga.exprs$CHD1[tcga.exprs$BCR == "NOBCR" & tcga.exprs$subtype == "ERG-"])$p.val,5)
p <- ggplot(tcga.exprs, aes(x=BCR, y=CHD1, fill=subtype)) + 
  geom_point(position = position_jitterdodge(.2)) + 
  geom_boxplot(outlier.shape = NA, alpha=0.4) + 
  ggtitle("CHD1 RNAseq (TCGA)") + 
  ylab("CHD1 Expression (z-score)") +
  xlab("Biochemical Relapse") +
  theme_bw() +
  ylim(c(-4,4)) +
  annotate("text", x = 1.5, y = 3.75, label = paste("Wilcoxon (ERG pos), p=",pval.ergpos, "; (ERG neg) p=",pval.ergneg, sep=""))
p

```

<a name="table1"/>

###  Table 1

Compile the table of clinical features about the patients separated based on if they relapsed (BCR) or if they did not relapse (NO BCR). t-test = independent t-test; WRS = Wilcoxon rank sum test; Fishers = Fisher’s exact test.
```{r, message=F, echo=F}
library(gridExtra)

### Table 1
#paste("BCR (n=", nrow(bcr), ")  No BCR (n=", nrow(nobcr), ")", sep="")
table1 <- data.frame(paste("(n=", nrow(bcr), ")", sep=""), paste("(n=", nrow(nobcr), ")", sep=""), paste(""))
names(table1) <- c("BCR", "No BCR", "p (test)")
row.names(table1)[nrow(table1)] <- "Sample Size"

#Age 
p <- round(t.test(bcr$AGE,nobcr$AGE)$p.val,2)
temp <- data.frame(paste(round(mean(bcr$AGE),1)," (",min(bcr$AGE), "-", max(bcr$AGE),")", sep=""), paste(round(mean(nobcr$AGE),1)," (",min(nobcr$AGE), "-", max(nobcr$AGE),")", sep=""), paste(p, " (t-test)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "Age"

#High Grade
p <- round(wilcox.test(bcr$Total.1,nobcr$Total.1)$p.val,2)
temp <- data.frame (paste(round(median(bcr$Total.1),1)," (",min(bcr$Total.1), "-", max(bcr$Total.1),")", sep=""), paste(round(median(nobcr$Total.1),1)," (",min(nobcr$Total.1), "-", max(nobcr$Total.1),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "High Grade"

#DOS (date of surgery)
p <- round(wilcox.test(bcr$DOS,nobcr$DOS)$p.val,2)
temp <- data.frame(paste(round(median(bcr$DOS),1)," (",min(bcr$DOS), "-", max(bcr$DOS),")", sep=""), paste(round(median(nobcr$DOS),1)," (",min(nobcr$DOS), "-", max(nobcr$DOS),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "Date of Surgery"


# Gleason
a <- c(length(bcr$Gleason[bcr$Gleason == "3+3 and 3+4"]), length(bcr$Gleason[bcr$Gleason == "4+3"]), length(bcr$Gleason[bcr$Gleason == "8+"]))
b <- c(length(nobcr$Gleason[nobcr$Gleason == "3+3 and 3+4"]), length(nobcr$Gleason[nobcr$Gleason == "4+3"]), length(nobcr$Gleason[nobcr$Gleason == "8+"]))
p <- round(chisq.test(matrix(c(a,b),nrow=length(a)))$p.val, 3)
temp <- data.frame(paste(a[1], " (",round(a[1]/sum(a),2)*100,"%)",sep=""), paste(b[1], " (",round(b[1]/sum(b),2)*100,"%)",sep=""), paste(p," (ChiSq)",sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "3+3 and 3+4 (%)"
temp <- data.frame(paste(a[2], " (",round(a[2]/sum(a),2)*100,"%)",sep=""), paste(b[2], " (",round(b[2]/sum(b),2)*100,"%)",sep=""), paste(""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "4+3 (%)"
temp <- data.frame(paste(a[3], " (",round(a[3]/sum(a),2)*100,"%)",sep=""), paste(b[3], " (",round(b[3]/sum(b),2)*100,"%)",sep=""), paste(""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "8+ (%)"

#PSA 
p <- round(wilcox.test(bcr$PSA,nobcr$PSA)$p.val,2)
temp <- data.frame(paste(round(median(bcr$PSA),1)," (",min(bcr$PSA), "-", max(bcr$PSA),")",sep=""), paste(round(median(nobcr$PSA),1)," (",min(nobcr$PSA), "-", max(nobcr$PSA),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "PSA"

#ERG
a <- length(hscore.1.bcr$HScoreERG[hscore.1.bcr$HScoreERG > 0])
b <- length(hscore.1.bcr$HScoreERG[hscore.1.bcr$HScoreERG == 0])
c <- length(hscore.1.nobcr$HScoreERG[hscore.1.nobcr$HScoreERG > 0])
d <- length(hscore.1.nobcr$HScoreERG[hscore.1.nobcr$HScoreERG == 0])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(b, " (", round(100*(b/(a+b)),1), "%)",sep=""), paste(d, " (", round(100*(d/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "ERG neg (%)"

# Cribriform
a <- length(hscore.1.bcr$Cribriform[hscore.1.bcr$Cribriform == 1])
b <- length(hscore.1.bcr$Cribriform[hscore.1.bcr$Cribriform == 0])
c <- length(hscore.1.nobcr$Cribriform[hscore.1.nobcr$Cribriform == 1])
d <- length(hscore.1.nobcr$Cribriform[hscore.1.nobcr$Cribriform == 0])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "Cribriform (%)"

# volume
p <- round(wilcox.test(bcr$Total,nobcr$Total)$p.val,2)
temp <- data.frame(paste(round(median(bcr$Total),1)," (",min(bcr$Total), "-", max(bcr$Total),")",sep=""), paste(round(median(nobcr$Total),1)," (",min(nobcr$Total), "-", max(nobcr$Total),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "Total Cancer Volume"

# PWEI
p <- round(wilcox.test(bcr$PWEI,nobcr$PWEI)$p.val,2)
temp <- data.frame(paste(round(median(bcr$PWEI),1)," (",min(bcr$PWEI), "-", max(bcr$PWEI),")",sep=""), paste(round(median(nobcr$PWEI),1)," (",min(nobcr$PWEI), "-", max(nobcr$PWEI),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "PWEI"

# Lymph Nodes
a <- length(bcr$NODES[bcr$NODES == "+"])
b <- length(bcr$NODES[bcr$NODES == "-"])
c <- length(nobcr$NODES[nobcr$NODES == "+"])
d <- length(nobcr$NODES[nobcr$NODES == "-"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "Lymph Nodes (%)"


# Seminal Vesicle
a <- length(bcr$SV[bcr$SV == "+"])
b <- length(bcr$SV[bcr$SV == "-"])
c <- length(nobcr$SV[nobcr$SV == "+"])
d <- length(nobcr$SV[nobcr$SV == "-"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "Seminal Vesicle (%)"

# Stage
bcr.tmp <- bcr$Clinical
bcr.tmp <- gsub("[ABCabc]", "",  bcr.tmp)
bcr.tmp <- gsub("T3", "T2",  bcr.tmp)
nobcr.tmp <- nobcr$Clinical
nobcr.tmp <- gsub("[ABCabc]", "",  nobcr.tmp)

a <- length(bcr.tmp[bcr.tmp == "T1"])
b <- length(bcr.tmp[bcr.tmp == "T2"])
c <- length(nobcr.tmp[nobcr.tmp == "T1"])
d <- length(nobcr.tmp[nobcr.tmp == "T2"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("BCR", "No BCR", "p (test)")
table1 <- rbind(table1,temp)
row.names(table1)[nrow(table1)] <- "Clinical Stage T1 vs. T2/T3 (%)"

# print the table
grid.table(table1)
```

<a name="table2"/>

###  Table 2

CHD1 and MAP3K7 H Scores for men with BCR and men without BCR.
```{r, message=F, echo=F}
# clean up the data to be in 1 matrix
hscore.combo <- rbind(hscore.1.bcr,hscore.1.nobcr)
hscore.combo$bcr <- c(rep("BCR",nrow(hscore.1.bcr)),rep("NOBCR", nrow(hscore.1.nobcr)))
hscore.combo$HScoreERG[hscore.combo$HScoreERG > 0] <- "ERGpos"
hscore.combo$HScoreERG[hscore.combo$HScoreERG == 0] <- "ERGneg"

cols.common <- intersect(colnames(bcr),colnames(nobcr))
bcr <- bcr[,cols.common]
nobcr <- nobcr[,cols.common]
bcr.all <- rbind(bcr,nobcr)
pats <- intersect(row.names(bcr.all), row.names(hscore.combo))
bcr.all <- bcr.all[pats,]
hscore.combo <- hscore.combo[pats,]
master.table <- cbind(bcr.all,hscore.combo)
master.table$bcr <- ifelse(grepl("NOBCR",master.table$bcr),0,1)
#print(paste("All men (n=",nrow(master.table),")",sep=""))

############################# Table 2A
############################# CHD1
# All men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1], master.table$HScoreCHD1[master.table$bcr==0])$p.val, 5)
table2a <- data.frame("BCR", nrow(master.table[master.table$bcr==1,]),round(mean(master.table$HScoreCHD1[master.table$bcr==1]),1),round(sd(master.table$HScoreCHD1[master.table$bcr==1]),1),paste(min(master.table$HScoreCHD1[master.table$bcr==1]), "-", max(master.table$HScoreCHD1[master.table$bcr==1]),sep=""),pval)
names(table2a) <- c("BCR Status", "N", "mean", "SD", "range", "p")
row.names(table2a)[nrow(table2a)] <- "CHD1, All Men"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0,]),round(mean(master.table$HScoreCHD1[master.table$bcr==0]),1),round(sd(master.table$HScoreCHD1[master.table$bcr==0]),1),paste(min(master.table$HScoreCHD1[master.table$bcr==0]), "-", max(master.table$HScoreCHD1[master.table$bcr==0]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- ""

# ERG pos men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"], master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGpos",]), round(mean(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"]), "-", max(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "CHD1, ERG+ Men"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGpos",]), round(mean(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"]), "-", max(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- " "

# ERG neg men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"], master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGneg",]), round(mean(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"]), "-", max(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "CHD1, ERG- Men"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGneg",]), round(mean(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"]), "-", max(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "  "


############################# MAP3K7
temp <- data.frame("","","","","","")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "-----"

# All men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1], master.table$HScoreMAP3K7[master.table$bcr==0])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1,]),round(mean(master.table$HScoreMAP3K7[master.table$bcr==1]),1),round(sd(master.table$HScoreMAP3K7[master.table$bcr==1]),1),paste(min(master.table$HScoreMAP3K7[master.table$bcr==1]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "MAP3K7, All Men"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0,]),round(mean(master.table$HScoreMAP3K7[master.table$bcr==0]),1),round(sd(master.table$HScoreMAP3K7[master.table$bcr==0]),1),paste(min(master.table$HScoreMAP3K7[master.table$bcr==0]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "   "

# ERG pos men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"], master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGpos",]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos"]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "MAP3K7, ERG+ Men"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGpos",]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos"]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "    "

# ERG neg men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"], master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGneg",]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg"]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "MAP3K7, ERG- Men"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGneg",]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg"]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2a <- rbind(table2a,temp)
row.names(table2a)[nrow(table2a)] <- "     "

############################# Table 2B
############################# CHD1 & Cribriform
# All men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 1], master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 1])$p.val, 5)
table2b <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$Cribriform == 1,]), round(mean(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 1]), "-", max(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 1]),sep=""),pval)
names(table2b) <- c("BCR Status", "N", "mean", "SD", "range", "p")
row.names(table2b)[nrow(table2b)] <- "CHD1, Men with Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$Cribriform == 1,]), round(mean(master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 1]),1),round(sd(master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 1]), "-", max(master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 1]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- ""

# ERG pos men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1], master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1,]), round(mean(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]), "-", max(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "CHD1, Men with ERG+ and Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1,]), round(mean(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]), "-", max(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- " "

# ERG neg men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1], master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1,]), round(mean(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]), "-", max(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "CHD1, Men with ERG- and Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1,]), round(mean(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]), "-", max(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "   "


############################# MAP3K7 & Cribriform
temp <- data.frame("","","","","","")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "-------"

# All men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$Cribriform == 1], master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 1])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$Cribriform == 1,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$Cribriform == 1]),1),paste(min(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$Cribriform == 1]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "MAP3K7, Men with Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$Cribriform == 1,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 1]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 1]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "    "

# ERG pos men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1], master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "MAP3K7, Men with ERG+ and Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "     "

# ERG neg men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1], master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "MAP3K7, Men with ERG- and Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2b <- rbind(table2b,temp)
row.names(table2b)[nrow(table2b)] <- "      "

############################# Table 2C
############################# CHD1 & No Cribriform
# All men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 0], master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 0])$p.val, 5)
table2c <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$Cribriform == 0,]), round(mean(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 0]), "-", max(master.table$HScoreCHD1[master.table$bcr==1 & master.table$Cribriform == 0]),sep=""),pval)
names(table2c) <- c("BCR Status", "N", "mean", "SD", "range", "p")
row.names(table2c)[nrow(table2c)] <- "CHD1, Men without Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$Cribriform == 0,]), round(mean(master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 0]),1),round(sd(master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 0]), "-", max(master.table$HScoreCHD1[master.table$bcr==0 & master.table$Cribriform == 0]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- ""

# ERG pos men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0], master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0,]), round(mean(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]), "-", max(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "CHD1, Men with ERG+ and No Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0,]), round(mean(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]), "-", max(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- " "

# ERG neg men
pval <- round(t.test(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0], master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0,]), round(mean(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]), "-", max(master.table$HScoreCHD1[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "CHD1, Men with ERG- and No Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0,]), round(mean(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]), "-", max(master.table$HScoreCHD1[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "   "


############################# MAP3K7 & No Cribriform
temp <- data.frame("","","","","","")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "-------"

# All men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$Cribriform == 0], master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 0])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$Cribriform == 0,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$Cribriform == 0]),1),paste(min(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$Cribriform == 0]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "MAP3K7, Men without Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$Cribriform == 0,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 0]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$Cribriform == 0]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "    "

# ERG pos men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0], master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "MAP3K7, Men with ERG+ and No Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "     "

# ERG neg men
pval <- round(t.test(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0], master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0])$p.val, 5)
temp <- data.frame("BCR", nrow(master.table[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==1 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),sep=""),pval)
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "MAP3K7, Men with ERG- and No Cribriform"

temp <- data.frame("No BCR", nrow(master.table[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0,]), round(mean(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),1), round(sd(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),1), paste(min(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]), "-", max(master.table$HScoreMAP3K7[master.table$bcr==0 & master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0]),sep=""),"")
names(temp) <- c("BCR Status", "N", "mean", "SD", "range", "p")
table2c <- rbind(table2c,temp)
row.names(table2c)[nrow(table2c)] <- "      "


# print out table
grid.table(table2a)
```

```{r, echo=F, message=F}
grid.table(table2b)
```

```{r, echo=F, message=F}
grid.table(table2c)
```

<a name="table3"/>

###  Table 3

Compile the table of clinical features about the patients separated based on ERG status. t-test = independent t-test; WRS = Wilcoxon rank sum test; Fishers = Fisher’s exact test.
```{r, message=F, echo=F}
### Table 3
#paste("ERG+ (n=", nrow(bcr), ")  ERG- (n=", nrow(nobcr), ")", sep="")
table3 <- data.frame(paste("(n=", length(master.table$HScoreERG[master.table$HScoreERG == "ERGpos"]), ")", sep=""), paste("(n=", length(master.table$HScoreERG[master.table$HScoreERG == "ERGneg"]), ")", sep=""), paste(""))
names(table3) <- c("ERG+", "ERG-", "p (test)")
row.names(table3)[nrow(table3)] <- "Sample Size"

#Age 
p <- round(t.test(master.table$AGE[master.table$HScoreERG == "ERGpos"],master.table$AGE[master.table$HScoreERG == "ERGneg"])$p.val,2)
temp <- data.frame(paste(round(mean(master.table$AGE[master.table$HScoreERG == "ERGpos"]),1)," (",min(master.table$AGE[master.table$HScoreERG == "ERGpos"]), "-", max(master.table$AGE[master.table$HScoreERG == "ERGpos"]),")", sep=""), paste(round(mean(master.table$AGE[master.table$HScoreERG == "ERGneg"]),1)," (",min(master.table$AGE[master.table$HScoreERG == "ERGneg"]), "-", max(master.table$AGE[master.table$HScoreERG == "ERGneg"]),")", sep=""), paste(p, " (t-test)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "Age"

#High Grade
p <- round(wilcox.test(master.table$Total.1[master.table$HScoreERG == "ERGpos"],master.table$Total.1[master.table$HScoreERG == "ERGneg"])$p.val,2)
temp <- data.frame (paste(round(median(master.table$Total.1[master.table$HScoreERG == "ERGpos"]),1)," (",min(master.table$Total.1[master.table$HScoreERG == "ERGpos"]), "-", max(master.table$Total.1[master.table$HScoreERG == "ERGpos"]),")", sep=""), paste(round(median(master.table$Total.1[master.table$HScoreERG == "ERGneg"]),1)," (",min(master.table$Total.1[master.table$HScoreERG == "ERGneg"]), "-", max(master.table$Total.1[master.table$HScoreERG == "ERGneg"]),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "High Grade"

#DOS (date of surgery)
p <- round(wilcox.test(master.table$DOS[master.table$HScoreERG == "ERGpos"],master.table$DOS[master.table$HScoreERG == "ERGneg"])$p.val,2)
temp <- data.frame(paste(round(median(master.table$DOS[master.table$HScoreERG == "ERGpos"]),1)," (",min(master.table$DOS[master.table$HScoreERG == "ERGpos"]), "-", max(master.table$DOS[master.table$HScoreERG == "ERGpos"]),")", sep=""), paste(round(median(master.table$DOS[master.table$HScoreERG == "ERGneg"]),1)," (",min(master.table$DOS[master.table$HScoreERG == "ERGneg"]), "-", max(master.table$DOS[master.table$HScoreERG == "ERGneg"]),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "Date of Surgery"

#PSA 
p <- round(wilcox.test(master.table$PSA[master.table$HScoreERG == "ERGpos"],master.table$PSA[master.table$HScoreERG == "ERGneg"])$p.val,2)
temp <- data.frame(paste(round(median(master.table$PSA[master.table$HScoreERG == "ERGpos"]),1)," (",min(master.table$PSA[master.table$HScoreERG == "ERGpos"]), "-", max(master.table$PSA[master.table$HScoreERG == "ERGpos"]),")",sep=""), paste(round(median(master.table$PSA[master.table$HScoreERG == "ERGneg"]),1)," (",min(master.table$PSA[master.table$HScoreERG == "ERGneg"]), "-", max(master.table$PSA[master.table$HScoreERG == "ERGneg"]),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "PSA"

# volume
p <- round(wilcox.test(master.table$Total[master.table$HScoreERG == "ERGpos"],master.table$Total[master.table$HScoreERG == "ERGneg"])$p.val,2)
temp <- data.frame(paste(round(median(master.table$Total[master.table$HScoreERG == "ERGpos"]),1)," (",min(master.table$Total[master.table$HScoreERG == "ERGpos"]), "-", max(master.table$Total[master.table$HScoreERG == "ERGpos"]),")",sep=""), paste(round(median(master.table$Total[master.table$HScoreERG == "ERGneg"]),1)," (",min(master.table$Total[master.table$HScoreERG == "ERGneg"]), "-", max(master.table$Total[master.table$HScoreERG == "ERGneg"]),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "Total Cancer Volume"

# PWEI
p <- round(wilcox.test(master.table$PWEI[master.table$HScoreERG == "ERGpos"],master.table$PWEI[master.table$HScoreERG == "ERGneg"])$p.val,2)
temp <- data.frame(paste(round(median(master.table$PWEI[master.table$HScoreERG == "ERGpos"]),1)," (",min(master.table$PWEI[master.table$HScoreERG == "ERGpos"]), "-", max(master.table$PWEI[master.table$HScoreERG == "ERGpos"]),")",sep=""), paste(round(median(master.table$PWEI[master.table$HScoreERG == "ERGneg"]),1)," (",min(master.table$PWEI[master.table$HScoreERG == "ERGneg"]), "-", max(master.table$PWEI[master.table$HScoreERG == "ERGneg"]),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "PWEI"

# Gleason
a <- c(length(master.table$Gleason[master.table$HScoreERG == "ERGpos" & master.table$Gleason == "3+3 and 3+4"]), length(master.table$Gleason[master.table$HScoreERG == "ERGpos" & master.table$Gleason == "4+3"]), length(master.table$Gleason[master.table$HScoreERG == "ERGpos" & master.table$Gleason == "8+"]))
b <- c(length(master.table$Gleason[master.table$HScoreERG == "ERGneg" & master.table$Gleason == "3+3 and 3+4"]), length(master.table$Gleason[master.table$HScoreERG == "ERGneg" & master.table$Gleason == "4+3"]), length(master.table$Gleason[master.table$HScoreERG == "ERGneg" & master.table$Gleason == "8+"]))
p <- round(chisq.test(matrix(c(a,b),nrow=length(a)))$p.val, 3)
temp <- data.frame(paste(a[1], " (",round(a[1]/sum(a),2)*100,"%)",sep=""), paste(b[1], " (",round(b[1]/sum(b),2)*100,"%)",sep=""), paste(p," (ChiSq)",sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "3+3 and 3+4 (%)"
temp <- data.frame(paste(a[2], " (",round(a[2]/sum(a),2)*100,"%)",sep=""), paste(b[2], " (",round(b[2]/sum(b),2)*100,"%)",sep=""), paste(""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "4+3 (%)"
temp <- data.frame(paste(a[3], " (",round(a[3]/sum(a),2)*100,"%)",sep=""), paste(b[3], " (",round(b[3]/sum(b),2)*100,"%)",sep=""), paste(""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "8+ (%)"

#BCR
a <- length(master.table$bcr[master.table$HScoreERG == "ERGpos" & master.table$bcr == 1])
b <- length(master.table$bcr[master.table$HScoreERG == "ERGpos" & master.table$bcr == 0])
c <- length(master.table$bcr[master.table$HScoreERG == "ERGneg" & master.table$bcr == 1])
d <- length(master.table$bcr[master.table$HScoreERG == "ERGneg" & master.table$bcr == 0])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(b, " (", round(100*(b/(a+b)),1), "%)",sep=""), paste(d, " (", round(100*(d/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "BCR (%)"

# Cribriform
a <- length(master.table$Cribriform[master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 1])
b <- length(master.table$Cribriform[master.table$HScoreERG == "ERGpos" & master.table$Cribriform == 0])
c <- length(master.table$Cribriform[master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 1])
d <- length(master.table$Cribriform[master.table$HScoreERG == "ERGneg" & master.table$Cribriform == 0])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "Cribriform (%)"

# Lymph Nodes
a <- length(master.table$NODES[master.table$HScoreERG == "ERGpos" & master.table$NODES == "+"])
b <- length(master.table$NODES[master.table$HScoreERG == "ERGpos" & master.table$NODES == "-"])
c <- length(master.table$NODES[master.table$HScoreERG == "ERGneg" & master.table$NODES == "+"])
d <- length(master.table$NODES[master.table$HScoreERG == "ERGneg" & master.table$NODES == "-"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "Lymph Nodes (%)"

# Seminal Vesicle
a <- length(master.table$SV[master.table$HScoreERG == "ERGpos" & master.table$SV == "+"])
b <- length(master.table$SV[master.table$HScoreERG == "ERGpos" & master.table$SV == "-"])
c <- length(master.table$SV[master.table$HScoreERG == "ERGneg" & master.table$SV == "+"])
d <- length(master.table$SV[master.table$HScoreERG == "ERGneg" & master.table$SV == "-"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "Seminal Vesicle (%)"

# Stage
tmp <- master.table$Clinical
tmp <- gsub("[ABCabc]", "",  tmp)
tmp <- gsub("T3", "T2",  tmp)

a <- length(tmp[master.table$HScoreERG == "ERGpos" & tmp == "T1"])
b <- length(tmp[master.table$HScoreERG == "ERGpos" & tmp == "T2"])
c <- length(tmp[master.table$HScoreERG == "ERGneg" & tmp == "T1"])
d <- length(tmp[master.table$HScoreERG == "ERGneg" & tmp == "T2"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ERG+", "ERG-", "p (test)")
table3 <- rbind(table3,temp)
row.names(table3)[nrow(table3)] <- "Clinical Stage T1 vs. T2/T3 (%)"

# print the table
grid.table(table3)
```

<a name="hscorerelapse"/>

### Compare Hscores in BCR patients

Compare the HScores for patients that relapsed to those that did not relapse plotted in different forms
```{r, echo=F, message=F}
#MAP3K7
p <- ggplot(hscore.combo, aes(x=bcr, y=HScoreMAP3K7)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(height=0, width=.3, cex=2) +
  theme_bw() + 
  ggtitle("MAP3K7") + 
  xlab("Biochemical Relapse") +
  stat_compare_means(method="wilcox.test")+
  ylab("HScore MAP3K7") +
  ylim(c(0,350))
p

pval.ergpos <- round(wilcox.test(hscore.combo$HScoreMAP3K7[hscore.combo$bcr == "BCR" & hscore.combo$HScoreERG == "ERGpos"], hscore.combo$HScoreMAP3K7[hscore.combo$bcr == "NOBCR" & hscore.combo$HScoreERG == "ERGpos"])$p.val,5)
pval.ergneg <- round(wilcox.test(hscore.combo$HScoreMAP3K7[hscore.combo$bcr == "BCR" & hscore.combo$HScoreERG == "ERGneg"], hscore.combo$HScoreMAP3K7[hscore.combo$bcr == "NOBCR" & hscore.combo$HScoreERG == "ERGneg"])$p.val,5)
p <- ggplot(hscore.combo, aes(x=bcr, y=HScoreMAP3K7, fill=HScoreERG)) + 
  geom_point(position = position_jitterdodge(.2)) + 
  geom_boxplot(outlier.shape = NA, alpha=0.4) + 
  ggtitle("MAP3K7") + 
  ylab("HScore MAP3K7") +
  xlab("Biochemical Relapse") +
  ylim(c(0,350)) +
  theme_bw() +
  annotate("text", x = 1.5, y = 335, label = paste("Wilcoxon (ERG pos), p=",pval.ergpos, "; (ERG neg) p=",pval.ergneg, sep=""))
p

#CHD1
p <- ggplot(hscore.combo, aes(x=bcr, y=HScoreCHD1)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(height=0, width=.3, cex=2) +
  theme_bw() + 
  ggtitle("CHD1") + 
  xlab("Biochemical Relapse") +
  stat_compare_means(method="wilcox.test")+
  ylab("HScore CHD1") +
  ylim(c(0,350))
p

pval.ergpos <- round(wilcox.test(hscore.combo$HScoreCHD1[hscore.combo$bcr == "BCR" & hscore.combo$HScoreERG == "ERGpos"], hscore.combo$HScoreCHD1[hscore.combo$bcr == "NOBCR" & hscore.combo$HScoreERG == "ERGpos"])$p.val,5)
pval.ergneg <- round(wilcox.test(hscore.combo$HScoreCHD1[hscore.combo$bcr == "BCR" & hscore.combo$HScoreERG == "ERGneg"], hscore.combo$HScoreCHD1[hscore.combo$bcr == "NOBCR" & hscore.combo$HScoreERG == "ERGneg"])$p.val,5)
p <- ggplot(hscore.combo, aes(x=bcr, y=HScoreCHD1, fill=HScoreERG)) + 
  geom_point(position = position_jitterdodge(.2)) + 
  geom_boxplot(outlier.shape = NA, alpha=0.4) + 
  ggtitle("CHD1") + 
  ylab("HScore") +
  xlab("") +
  ylim(c(0,350)) +
  theme_bw() +
  annotate("text", x = 1.5, y = 335, label = paste("Wilcoxon (ERG pos), p=",pval.ergpos, "; (ERG neg) p=",pval.ergneg, sep=""))
p


plot(hscore.combo$HScoreMAP3K7, hscore.combo$HScoreCHD1, col="white", main="HScore of MAP3K7 compared to CHD1", ylab="HScore CHD1", xlab="HScore MAP3K7", ylim=c(0,400), xlim=c(0,300))
points(hscore.combo$HScoreMAP3K7[hscore.combo$bcr=="BCR" & hscore.combo$HScoreERG=="ERGpos"],hscore.combo$HScoreCHD1[hscore.combo$bcr=="BCR" & hscore.combo$HScoreERG=="ERGpos"], col="red", pch=15)
points(hscore.combo$HScoreMAP3K7[hscore.combo$bcr=="BCR" & hscore.combo$HScoreERG=="ERGneg"],hscore.combo$HScoreCHD1[hscore.combo$bcr=="BCR" & hscore.combo$HScoreERG=="ERGneg"], col="red", pch=18)
points(hscore.combo$HScoreMAP3K7[hscore.combo$bcr=="NOBCR" & hscore.combo$HScoreERG=="ERGpos"],hscore.combo$HScoreCHD1[hscore.combo$bcr=="NOBCR" & hscore.combo$HScoreERG=="ERGpos"], col="blue", pch=15)
points(hscore.combo$HScoreMAP3K7[hscore.combo$bcr=="NOBCR" & hscore.combo$HScoreERG=="ERGneg"],hscore.combo$HScoreCHD1[hscore.combo$bcr=="NOBCR" & hscore.combo$HScoreERG=="ERGneg"], col="blue", pch=18)
legend("topleft", legend=c("BCR, ERG+", "BCR, ERG-", "NO BCR, ERG+", "NO BCR, ERG-"), col=c("red","red", "blue", "blue"), pch=c(15,18,15,18))
```
<a name="nc4"/>

### Compare Hscores in BCR patients for patients with ICC and NC4

Compare the HScores and BCR for ICC and NC4 patients 
```{r, echo=F, message=F, warning=F}
master.table.nc4 <- master.table[master.table$Gleason != "3+3 and 3+4",]
icc <- master.table.nc4[master.table.nc4$Cribriform == 1,]
nc4 <- master.table.nc4[master.table.nc4$Cribriform == 0,]

table.nc4 <- data.frame(paste("(n=", nrow(icc), ")", sep=""), paste("(n=", nrow(nc4), ")", sep=""), paste(""))
names(table.nc4) <- c("ICC", "NC4", "p (test)")
row.names(table.nc4)[nrow(table.nc4)] <- "Sample Size"

#Age 
p <- round(t.test(icc$AGE,nc4$AGE)$p.val,2)
temp <- data.frame(paste(round(mean(icc$AGE),1)," (",min(icc$AGE), "-", max(icc$AGE),")", sep=""), paste(round(mean(nc4$AGE),1)," (",min(nc4$AGE), "-", max(nc4$AGE),")", sep=""), paste(p, " (t-test)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "Age"

#High Grade
p <- round(wilcox.test(icc$Total.1,nc4$Total.1)$p.val,2)
temp <- data.frame (paste(round(median(icc$Total.1),1)," (",min(icc$Total.1), "-", max(icc$Total.1),")", sep=""), paste(round(median(nc4$Total.1),1)," (",min(nc4$Total.1), "-", max(nc4$Total.1),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "High Grade"

#DOS (date of surgery)
p <- round(wilcox.test(icc$DOS,nc4$DOS)$p.val,2)
temp <- data.frame(paste(round(median(icc$DOS),1)," (",min(icc$DOS), "-", max(icc$DOS),")", sep=""), paste(round(median(nc4$DOS),1)," (",min(nc4$DOS), "-", max(nc4$DOS),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "Date of Surgery"


# Gleason
a <- c(length(icc$Gleason[icc$Gleason == "4+3"]), length(icc$Gleason[icc$Gleason == "8+"]))
b <- c(length(nc4$Gleason[nc4$Gleason == "4+3"]), length(nc4$Gleason[nc4$Gleason == "8+"]))
p <- round(chisq.test(matrix(c(a,b),nrow=length(a)))$p.val, 3)
temp <- data.frame(paste(a[1], " (",round(a[1]/sum(a),2)*100,"%)",sep=""), paste(b[1], " (",round(b[1]/sum(b),2)*100,"%)",sep=""), paste(p," (ChiSq)",sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "4+3 (%)"
temp <- data.frame(paste(a[2], " (",round(a[2]/sum(a),2)*100,"%)",sep=""), paste(b[2], " (",round(b[2]/sum(b),2)*100,"%)",sep=""), paste(""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "8+ (%)"

#PSA 
p <- round(wilcox.test(icc$PSA,nc4$PSA)$p.val,2)
temp <- data.frame(paste(round(median(icc$PSA),1)," (",min(icc$PSA), "-", max(icc$PSA),")",sep=""), paste(round(median(nc4$PSA),1)," (",min(nc4$PSA), "-", max(nc4$PSA),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "PSA"

#ERG
a <- length(icc$HScoreERG[icc$HScoreERG == "ERGpos"])
b <- length(icc$HScoreERG[icc$HScoreERG == "ERGneg"])
c <- length(nc4$HScoreERG[nc4$HScoreERG == "ERGpos"])
d <- length(nc4$HScoreERG[nc4$HScoreERG == "ERGneg"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(b, " (", round(100*(b/(a+b)),1), "%)",sep=""), paste(d, " (", round(100*(d/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "ERG neg (%)"

# BCR
a <- length(icc$bcr[icc$bcr == 1])
b <- length(icc$bcr[icc$bcr == 0])
c <- length(nc4$bcr[nc4$bcr == 1])
d <- length(nc4$bcr[nc4$bcr == 0])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "BCR (%)"

# volume
p <- round(wilcox.test(icc$Total,nc4$Total)$p.val,2)
temp <- data.frame(paste(round(median(icc$Total),1)," (",min(icc$Total), "-", max(icc$Total),")",sep=""), paste(round(median(nc4$Total),1)," (",min(nc4$Total), "-", max(nc4$Total),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "Total Cancer Volume"

# PWEI
p <- round(wilcox.test(icc$PWEI,nc4$PWEI)$p.val,2)
temp <- data.frame(paste(round(median(icc$PWEI),1)," (",min(icc$PWEI), "-", max(icc$PWEI),")",sep=""), paste(round(median(nc4$PWEI),1)," (",min(nc4$PWEI), "-", max(nc4$PWEI),")", sep=""), paste(p, " (WRS)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "PWEI"

# Lymph Nodes
a <- length(icc$NODES[icc$NODES == "+"])
b <- length(icc$NODES[icc$NODES == "-"])
c <- length(nc4$NODES[nc4$NODES == "+"])
d <- length(nc4$NODES[nc4$NODES == "-"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "Lymph Nodes (%)"


# Seminal Vesicle
a <- length(icc$SV[icc$SV == "+"])
b <- length(icc$SV[icc$SV == "-"])
c <- length(nc4$SV[nc4$SV == "+"])
d <- length(nc4$SV[nc4$SV == "-"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "Seminal Vesicle (%)"

# Stage
icc.tmp <- icc$Clinical
icc.tmp <- gsub("[ABCabc]", "",  icc.tmp)
icc.tmp <- gsub("T3", "T2",  icc.tmp)
nc4.tmp <- nc4$Clinical
nc4.tmp <- gsub("[ABCabc]", "",  nc4.tmp)

a <- length(icc.tmp[icc.tmp == "T1"])
b <- length(icc.tmp[icc.tmp == "T2"])
c <- length(nc4.tmp[nc4.tmp == "T1"])
d <- length(nc4.tmp[nc4.tmp == "T2"])
m <- matrix(c(a,b,c,d), nrow=2)
p <- round(fisher.test(m)$p.val,3)
temp <- data.frame(paste(a, " (", round(100*(a/(a+b)),1), "%)",sep=""), paste(c, " (", round(100*(c/(c+d)),1),"%)",sep=""), paste(p, " (Fishers)", sep=""))
names(temp) <- c("ICC", "NC4", "p (test)")
table.nc4 <- rbind(table.nc4,temp)
row.names(table.nc4)[nrow(table.nc4)] <- "Clinical Stage T1 vs. T2/T3 (%)"

# print the table
grid.table(table.nc4)


############################# plots
#MAP3K7
master.table.nc4$Cribriform <- ifelse(master.table.nc4$Cribriform == 1,"Cribriform","No Cribriform")
p <- ggplot(master.table.nc4, aes(x=Cribriform, y=HScoreMAP3K7)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(height=0, width=.3, cex=2) +
  theme_bw() + 
  ggtitle("MAP3K7, ICC and NC4 patients") + 
  xlab("Biochemical Relapse") +
  stat_compare_means(method="wilcox.test")+
  ylab("HScore MAP3K7") +
  ylim(c(0,350))
p

pval.ergpos <- round(wilcox.test(master.table.nc4$HScoreMAP3K7[master.table.nc4$Cribriform == "Cribriform" & master.table.nc4$HScoreERG == "ERGpos"], master.table.nc4$HScoreMAP3K7[master.table.nc4$Cribriform == "No Cribriform" & master.table.nc4$HScoreERG == "ERGpos"])$p.val,5)
pval.ergneg <- round(wilcox.test(master.table.nc4$HScoreMAP3K7[master.table.nc4$Cribriform == "Cribriform" & master.table.nc4$HScoreERG == "ERGneg"], master.table.nc4$HScoreMAP3K7[master.table.nc4$Cribriform == "No Cribriform" & master.table.nc4$HScoreERG == "ERGneg"])$p.val,5)
p <- ggplot(master.table.nc4, aes(x=Cribriform, y=HScoreMAP3K7, fill=HScoreERG)) + 
  geom_point(position = position_jitterdodge(.2)) + 
  geom_boxplot(outlier.shape = NA, alpha=0.4) + 
  ggtitle("MAP3K7, ICC vs. NC4") + 
  ylab("HScore MAP3K7") +
  xlab("Biochemical Relapse") +
  ylim(c(0,350)) +
  theme_bw() +
  annotate("text", x = 1.5, y = 335, label = paste("Wilcoxon (ERG pos), p=",pval.ergpos, "; (ERG neg) p=",pval.ergneg, sep=""))
p

#CHD1
p <- ggplot(master.table.nc4, aes(x=Cribriform, y=HScoreCHD1)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(height=0, width=.3, cex=2) +
  theme_bw() + 
  ggtitle("CHD1") + 
  xlab("Biochemical Relapse") +
  stat_compare_means(method="wilcox.test")+
  ylab("HScore CHD1") +
  ylim(c(0,350))
p

pval.ergpos <- round(wilcox.test(master.table.nc4$HScoreCHD1[master.table.nc4$Cribriform == "Cribriform" & master.table.nc4$HScoreERG == "ERGpos"], master.table.nc4$HScoreCHD1[master.table.nc4$Cribriform == "No Cribriform" & master.table.nc4$HScoreERG == "ERGpos"])$p.val,5)
pval.ergneg <- round(wilcox.test(master.table.nc4$HScoreCHD1[master.table.nc4$Cribriform == "Cribriform" & master.table.nc4$HScoreERG == "ERGneg"], master.table.nc4$HScoreCHD1[master.table.nc4$Cribriform == "No Cribriform" & master.table.nc4$HScoreERG == "ERGneg"])$p.val,5)
p <- ggplot(master.table.nc4, aes(x=Cribriform, y=HScoreCHD1, fill=HScoreERG)) + 
  geom_point(position = position_jitterdodge(.2)) + 
  geom_boxplot(outlier.shape = NA, alpha=0.4) + 
  ggtitle("CHD1, ICC vs. NC4") + 
  ylab("HScore") +
  xlab("") +
  ylim(c(0,350)) +
  theme_bw() +
  annotate("text", x = 1.5, y = 335, label = paste("Wilcoxon (ERG pos), p=",pval.ergpos, "; (ERG neg) p=",pval.ergneg, sep=""))
p

```

<a name="univariate"/>

### Table 4

Univariate logistic regression analyses modeling the risk for BCR compared to no BCR
```{r, message=F, echo=F, warning=F}
rocplotvalues <- function(tmpmodel, tmpdata) {
  prec <- 0
  recall <- 0
  fpr <- 0
  acc <- 0
  for(i in seq(0,1,0.01)) {
    tmpdata.pr <- predict.glm(tmpmodel,tmpdata, type="response") > i
    tmpdata.pr.F <- tmpdata$bcr[tmpdata.pr==F]
    tmpdata.pr.T <- tmpdata$bcr[tmpdata.pr==T]
    tp <- length(tmpdata.pr.F[tmpdata.pr.F==1])
    fp <- length(tmpdata.pr.F[tmpdata.pr.F==0])
    fn <- length(tmpdata.pr.T[tmpdata.pr.T==1])
    tn <- length(tmpdata.pr.T[tmpdata.pr.T==0])
    prec <- c(prec,round(tp / (tp+fp),3))
    recall <- c(recall, round(tp /(tp+fn),3))
    fpr <- c(fpr,round(fp/(fp+tn),3))
    acc <- c(acc,round((tp+fp)/(tp+fp+tn+fn),3))
  }
  prec <- prec[2:length(prec)]
  recall <- recall[2:length(recall)]
  fpr <- fpr[2:length(fpr)]
  acc <- acc[2:length(acc)]
  return(data.frame(prec=prec,recall=recall,fpr=fpr,acc=acc))
}

simple_auc <- function(TPR, FPR){
  dFPR <- c(diff(FPR), 0)
  dTPR <- c(diff(TPR), 0)
  sum(TPR * dFPR) + sum(dTPR * dFPR)/2
}

##### univariate logistic modeling 

#Age
model <- glm(bcr ~ AGE, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
univariatematrix <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model))[2],3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3), auc)
names(univariatematrix) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
row.names(univariatematrix)[nrow(univariatematrix)] <- "Age"

#DOS
model <- glm(bcr ~ DOS, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3), auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "DOS"

#High Grade
model <- glm(bcr ~ Total.1, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3), auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "High Grade"

#PSA
model <- glm(bcr ~ PSA, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "PSA"


#Volume
model <- glm(bcr ~ Total, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "Total Cancer Volume"

#PWEI
model <- glm(bcr ~ PWEI, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "PWEI"

#ERG
tmp <- data.frame(HScoreERG=master.table$HScoreERG,bcr=master.table$bcr)
tmp$HScoreERG <- ifelse(grepl("ERGpos",tmp$HScoreERG),1,0)
model <- glm(bcr ~ HScoreERG, data=tmp, family=binomial(link="logit"))
rp <- rocplotvalues(model, tmp)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "ERG pos vs. neg"

#Cribriform
model <- glm(bcr ~ Cribriform, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "Cribriform present vs. not"

#Gleason
master.table <- cbind(master.table, GleasonLow=master.table$Gleason, GleasonHigh=master.table$Gleason)
master.table$GleasonLow <- ifelse(grepl("3\\+3 and 3\\+4",master.table$GleasonLow),1,0)
master.table$GleasonHigh <- ifelse(grepl("4\\+3",master.table$GleasonHigh),1,0)
model <- glm(bcr ~ GleasonLow, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "Gleason 3+3 and 3+4"
model <- glm(bcr ~ GleasonHigh, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "Gleason 4+3"

# Lymph Nodes
pats <- row.names(master.table)[master.table$NODES != '']
tmp <- master.table[pats,]
tmp$NODES <- ifelse(grepl("\\+",tmp$NODES),1,0)
model <- glm(bcr ~ NODES, data=tmp, family=binomial(link="logit"))
rp <- rocplotvalues(model, tmp)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(tmp$bcr), nrow(tmp)-sum(tmp$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
#univariatematrix <- rbind(univariatematrix,temp)
#row.names(univariatematrix)[nrow(univariatematrix)] <- "Lymph Node + vs. -"

# Seminal Vesicle
pats <- row.names(master.table)[master.table$SV != '']
tmp <- master.table[pats,]
tmp$SV <- ifelse(grepl("\\+",tmp$SV),1,0)
model <- glm(bcr ~ SV, data=tmp, family=binomial(link="logit"))
rp <- rocplotvalues(model, tmp)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(tmp$bcr), nrow(tmp)-sum(tmp$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
#univariatematrix <- rbind(univariatematrix,temp)
#row.names(univariatematrix)[nrow(univariatematrix)] <- "Seminal Vesicle + vs. -"

# Stage
master.table <- cbind(master.table, ClinicalT1 = master.table$Clinical, ClinicalT2 = master.table$Clinical)
master.table$ClinicalT1 <- ifelse(grepl("T1",master.table$ClinicalT1),1,0)
master.table$ClinicalT2 <- gsub("T3", "T2",  master.table$ClinicalT2)
master.table$ClinicalT2 <- ifelse(grepl("T2",master.table$ClinicalT2),1,0)
model <- glm(bcr ~ ClinicalT1, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "Clinical Stage T1 vs not"
model <- glm(bcr ~ ClinicalT2, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "Clinical Stage T2 vs not"

#MAP3K7
model <- glm(bcr ~ HScoreMAP3K7, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "HScore MAP3K7"
master.table <- cbind(master.table, HScoreMAP3K7split=master.table$HScoreMAP3K7)
master.table$HScoreMAP3K7split <- ifelse(master.table$HScoreMAP3K7split < 150,1,0)
model <- glm(bcr ~ HScoreMAP3K7split, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "MAP3K7 HScore < 150 vs 150+"

#CHD1
model <- glm(bcr ~ HScoreCHD1, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "HScore CHD1"
master.table <- cbind(master.table, HScoreCHD1split=master.table$HScoreCHD1)
master.table$HScoreCHD1split <- ifelse(master.table$HScoreCHD1split < 150,1,0)
model <- glm(bcr ~ HScoreCHD1split, data=master.table, family=binomial(link="logit"))
rp <- rocplotvalues(model, master.table)
auc <- round(simple_auc(rp$fpr,rp$recall),3)
temp <- data.frame(sum(master.table$bcr), nrow(master.table)-sum(master.table$bcr), round(exp(coef(model)[2]),3), paste(round(exp(confint(model, level=0.95))[2],3), ",", round(exp(confint(model, level=0.95))[4],3), sep=""), round(coef(summary(model))[2,4],3),auc)
names(temp) <- c("BCR", "No BCR", "OR", "CI (95%)", "p", "AUC")
univariatematrix <- rbind(univariatematrix,temp)
row.names(univariatematrix)[nrow(univariatematrix)] <- "CHD1 HScore < 150 vs 150+"

#print the matrix
grid.table(univariatematrix)

```

<a name="multivariate"/>

### Table 5

Multivariate logistic regression analyses modeling BCR status as the output variable.  The modeling starts with HScores of CHD1 and MAP3K7, and then adds individual variables to determing if they add or subtract from the overall model performance.  
```{r, echo=F, message=F}
# try all variables
cols <- c("DOS","AGE","PSA","Vol","Total","Total.1","GleasonHigh", "GleasonLow","Cribriform", "HScoreMAP3K7split", "HScoreCHD1split", "bcr")
master.table.multi <- master.table[,cols]

model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
#multivariatematrix <- data.frame(OR = round(exp(coef(model)),3), CI = paste(round(exp(confint(model, level=0.95))[,1],3), ",", round(exp(confint(model, level=0.95))[,2],3), sep=""), p = round(coef(summary(model))[,4],3))
#print(paste(cols))
#grid.table(multivariatematrix)
#model.eval <- rocplotvalues(model, master.table.multi)
#auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
#plot(model.eval$recall, model.eval$fpr, type="l", ylim=c(0,1), xlim=c(0,1), ylab="True Positive Rate",xlab="False Positive Rate", main=paste("ROC plot for logistic regression, AUC=",auc))


#filtered set of variables
cols <- c("HScoreMAP3K7", "HScoreCHD1", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
#plot(model.eval$recall, model.eval$fpr, type="l", ylim=c(0,1), xlim=c(0,1), ylab="True Positive Rate",xlab="False Positive Rate", main=paste("ROC plot for logistic regression, AUC=",auc))
#multivariatematrix <- data.frame(OR = round(exp(coef(model)),3), CI = paste(round(exp(confint(model, level=0.95))[,1],3), ",", round(exp(confint(model, level=0.95))[,2],3), sep=""), p = round(coef(summary(model))[,4],3))
#print(paste(cols))
#grid.table(multivariatematrix)

##################################

### Multivariate using Hscores of MAP3K7 as the base model

#establish the baseline model of just HScores of MAP3K7
cols <- c("HScoreMAP3K7", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc.base <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
modelcontribution <- data.frame(AUC=auc.base, Diff=0)
names(modelcontribution) <- c("AUC", "Difference")
row.names(modelcontribution)[nrow(modelcontribution)] <- "MAP3K7 HScores"

# test the contribution of other variables to complement the MAP3K7 model
#Hscore CHD1
cols <- c("HScoreMAP3K7", "HScoreCHD1", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+CHD1 HScore"

#Gleason High
cols <- c("HScoreMAP3K7", "GleasonHigh","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Gleason High"

#Gleason Low
cols <- c("HScoreMAP3K7", "GleasonLow","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Gleason Low"

#Age
cols <- c("HScoreMAP3K7", "AGE","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Age"

#PSA
cols <- c("HScoreMAP3K7", "PSA","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ PSA"

#DOS
cols <- c("HScoreMAP3K7", "DOS","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Date of Surgery"

#Cancer Volume
cols <- c("HScoreMAP3K7", "Total","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Total Cancer Volume"

# %High Grade 
cols <- c("HScoreMAP3K7", "Total.1","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ High Grade"

#Prostate Weight PWEI
cols <- c("HScoreMAP3K7", "PWEI","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ PWEI"

#Clinical Stage >T2
cols <- c("HScoreMAP3K7", "ClinicalT2","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Clinical > T2"

#Clincal Stage T1
cols <- c("HScoreMAP3K7",  "ClinicalT1","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Clinical T1"

#Cribriform presence
cols <- c("HScoreMAP3K7","Cribriform","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Cribriform"

#ERG status
cols <- c("HScoreMAP3K7", "HScoreERG","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ ERG status"

#SV
cols <- c("HScoreMAP3K7", "SV","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Seminal Vesicle"

#Lymph Nodes
cols <- c("HScoreMAP3K7", "NODES","bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3) 
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ Lymph Nodes"


grid.table(modelcontribution)


#############
### Multivariate using an optimized set of post surgery clincal features as the base


# establish the baseline model of optimized combination of post surgery variables
cols <- c("Total","PSA", "SV","NODES", "GleasonLow", "GleasonHigh", "AGE", "DOS", "PWEI", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc.base <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
modelcontribution <- data.frame(AUC=auc.base, Diff=0)
names(modelcontribution) <- c("AUC", "Difference")
row.names(modelcontribution)[nrow(modelcontribution)] <- "Optimized clinical factors"


# Add map3k7
cols <- c("Total","PSA", "SV", "NODES", "GleasonLow", "GleasonHigh", "AGE", "DOS", "PWEI", "HScoreMAP3K7", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ MAP3K7 HScores"

# Add CHD1
cols <- c("Total","PSA","SV", "NODES", "GleasonLow", "GleasonHigh", "AGE", "DOS", "PWEI" , "HScoreCHD1", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ CHD1 Hscores"

# Add MAP3K7 and CHD1
cols <- c("Total","PSA","SV", "NODES", "GleasonLow", "GleasonHigh", "AGE", "DOS", "PWEI", "HScoreMAP3K7", "HScoreCHD1", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ MAP3K7 & CHD1 Hscores"


#############
### Multivariate using an CAPRA-S score as the base model
# establish the baseline model of CAPRA.S score
cols <- c("CAPRA.S", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc.base <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
modelcontribution <- data.frame(AUC=auc.base, Diff=0)
names(modelcontribution) <- c("AUC", "Difference")
row.names(modelcontribution)[nrow(modelcontribution)] <- "CAPRA.S Base"

# Add map3k7
cols <- c("CAPRA.S", "HScoreMAP3K7", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ MAP3K7"

# Add CHD1
cols <- c("CAPRA.S", "HScoreCHD1", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ CHD1"

# Add MAP3K7 and CHD1
cols <- c("CAPRA.S", "HScoreMAP3K7", "HScoreCHD1", "bcr")
master.table.multi <- master.table[,cols]
model <- glm(bcr ~ ., data=master.table.multi, family=binomial(link="logit"))
model.eval <- rocplotvalues(model, master.table.multi)
auc <- round(simple_auc(model.eval$fpr,model.eval$recall),3)
diff <- round(auc - auc.base,3)
temp <- data.frame(AUC=auc, Diff=diff)
names(temp) <- c("AUC", "Difference")
modelcontribution <- rbind(modelcontribution,temp)
row.names(modelcontribution)[nrow(modelcontribution)] <- "+ MAP3K7&CHD1"

```

<a name="predictor"/>

### Decision boundry for predicting BCR using MAP3K7 and CHD1 HScores.

Use logistic regression to plot the decision boundry. The section below will iterate over random sampling of the input data. There are many samples that were Hscored more than once, so the program will randomly select one of the sets of Hscores for a single patient.  We identify the prediction threhsold that returns a 0.25 recall over 100 iterations of the model. The final plot is previously defined set of Hscores to use (established by Torkko) using the established prediction threshold. 
```{r, message=F, echo=F}
library(caret)

decisionplot <- function(model, data, class = NULL, predict_type = "class",
                         resolution = 100, showgrid = TRUE, ...) {
  
  if(!is.null(class)) cl <- data[,class] else cl <- 1
  data <- data[,1:2]
  pl <- cl
  cl <-  ifelse(grepl("NOBCR",cl),4,2)
  pl <-  ifelse(grepl("NOBCR",pl),19,19)
  k <- length(unique(cl))
  
  plot(data, col = as.integer(cl), pch = as.integer(pl), ...)
  
  # make grid
  r <- sapply(data, range, na.rm = TRUE)
  xs <- seq(r[1,1], r[2,1], length.out = resolution)
  ys <- seq(r[1,2], r[2,2], length.out = resolution)
  g <- cbind(rep(xs, each=resolution), rep(ys, time = resolution))
  colnames(g) <- colnames(r)
  g <- as.data.frame(g)
  
  ### guess how to get class labels from predict
  ### (unfortunately not very consistent between models)
  p <- predict(model, g, type = predict_type)
  if(is.list(p)) p <- p$class
  p <- as.factor(p)
  
  if(showgrid) points(g, col = as.integer(p)+1L, pch = ".")
  
  z <- matrix(as.integer(p), nrow = resolution, byrow = TRUE)
  contour(xs, ys, z, add = TRUE, drawlabels = FALSE,
          lwd = 2, levels = (1:(k-1))+.5)
  legend("bottomright", legend=c("BCR", "NO BCR"), col=c("red", "blue"), pch=c(19,19))
  
  invisible(z)
}


rocplotvalues <- function(tmpmodel, tmpdata) {
  prec <- 0
  recall <- 0
  fpr <- 0
  acc <- 0
  for(i in seq(0,1,0.01)) {
    tmpdata.pr <- predict.glm(tmpmodel,tmpdata, type="response") > i
    tmpdata.pr.F <- tmpdata$BCR[tmpdata.pr==F]
    tmpdata.pr.T <- tmpdata$BCR[tmpdata.pr==T]
    tp <- length(tmpdata.pr.F[tmpdata.pr.F=="BCR"])
    fp <- length(tmpdata.pr.F[tmpdata.pr.F=="NOBCR"])
    fn <- length(tmpdata.pr.T[tmpdata.pr.T=="BCR"])
    tn <- length(tmpdata.pr.T[tmpdata.pr.T=="NOBCR"])
    prec <- c(prec,round(tp / (tp+fp),3))
    recall <- c(recall, round(tp /(tp+fn),3))
    fpr <- c(fpr,round(fp/(fp+tn),3))
    acc <- c(acc,round((tp+fp)/(tp+fp+tn+fn),3))
  }
  prec <- prec[2:length(prec)]
  recall <- recall[2:length(recall)]
  fpr <- fpr[2:length(fpr)]
  acc <- acc[2:length(acc)]
  return(data.frame(prec=prec,recall=recall,fpr=fpr,acc=acc))
}

################################3
# sample the data to determine the threshold to set the prediction limit 0.25 recall
thres <- array()
pats.uni <- unique(hscores$ID)
recall.thres = 0.25
for (q in 1:100){
  df <- data.frame(HScoreMAP3K7=integer(), HScoreCHD1=integer())
  for(i in 1:length(pats.uni)) {
    tmpmat <- hscores[hscores$ID == pats.uni[i],]
    j <- sample(nrow(tmpmat),1)
    df <- rbind(df, data.frame(HScoreMAP3K7=tmpmat$HScoreMAP3K7[j], HScoreCHD1=tmpmat$HScoreCHD1[j]))
    row.names(df)[nrow(df)] <- as.character(tmpmat$ID[j])
  }
  # remove any samples with missing Hscores for ERG, CHD1 or MAP3K7
  df <- df[!is.na(df$HScoreCHD1),]
  df <- df[!is.na(df$HScoreMAP3K7),]
  pats <- intersect(row.names(hscore.combo), row.names(df))
  df <- df[pats,]
  tmp.bcr <- hscore.combo[pats,]
  df <- cbind(df, BCR=tmp.bcr$bcr)
  model <- glm(BCR ~ ., data = df, family=binomial(link='logit'))
  class(model) <- c("lr", class(model))

  model.eval <- rocplotvalues(model, df)
  min.recall <- min(model.eval$recall[model.eval$recall > recall.thres])
  k <- match(min.recall, model.eval$recall)
  thres <- c(thres,seq(0,1,0.01)[k])
}
thres <- thres[2:length(thres)]

plot(density(thres), main=paste("Distribution of prediction thresholds at ",recall.thres," recall\nThe median values=",median(thres),sep=""), xlab="Prediction Threshold")

# build predictor using a logistic regression
tmp <- data.frame("MAP3K7 HScore"=as.numeric(master.table$HScoreMAP3K7), "CHD1 HScore"=as.numeric(master.table$HScoreCHD1), BCR=factor(hscore.combo$bcr))
model <- glm(BCR ~ ., data = tmp, family=binomial(link='logit'))
class(model) <- c("lr", class(model))
predict.lr <- function(object, newdata, ...)
  predict.glm(object, newdata, type = "response") > median(thres)

tmp.pr <- predict.lr(model,tmp)
tmp.pr.F <- tmp$BCR[tmp.pr==F]
tmp.pr.T <- tmp$BCR[tmp.pr==T]
tp <- length(tmp.pr.F[tmp.pr.F=="BCR"])
fp <- length(tmp.pr.F[tmp.pr.F=="NOBCR"])
fn <- length(tmp.pr.T[tmp.pr.T=="BCR"])
tn <- length(tmp.pr.T[tmp.pr.T=="NOBCR"])
prec <- round(tp / (tp+fp),3)
recall <- round(tp /(tp+fn),3)
acc <- round((tp+tn)/(tp+tn+fp+fn),3)

decisionplot(model, tmp, class = "BCR", main = paste("Logistic Regression, Acc=",acc, ", Prec=",prec, " @ ",recall.thres, " Recall", sep=""), resolution=1000, showgrid=F)




######################## 
# sample the data to determine the threshold to set the prediction limit 0.5 recall
thres <- array()
pats.uni <- unique(hscores$ID)
recall.thres = 0.5
for (q in 1:100){
  df <- data.frame(HScoreMAP3K7=integer(), HScoreCHD1=integer())
  for(i in 1:length(pats.uni)) {
    tmpmat <- hscores[hscores$ID == pats.uni[i],]
    j <- sample(nrow(tmpmat),1)
    df <- rbind(df, data.frame(HScoreMAP3K7=tmpmat$HScoreMAP3K7[j], HScoreCHD1=tmpmat$HScoreCHD1[j]))
    row.names(df)[nrow(df)] <- as.character(tmpmat$ID[j])
  }
  # remove any samples with missing Hscores for ERG, CHD1 or MAP3K7
  df <- df[!is.na(df$HScoreCHD1),]
  df <- df[!is.na(df$HScoreMAP3K7),]
  pats <- intersect(row.names(hscore.combo), row.names(df))
  df <- df[pats,]
  tmp.bcr <- hscore.combo[pats,]
  df <- cbind(df, BCR=tmp.bcr$bcr)
  model <- glm(BCR ~ ., data = df, family=binomial(link='logit'))
  class(model) <- c("lr", class(model))

  model.eval <- rocplotvalues(model, df)
  min.recall <- min(model.eval$recall[model.eval$recall > recall.thres])
  k <- match(min.recall, model.eval$recall)
  thres <- c(thres,seq(0,1,0.01)[k])
}
thres <- thres[2:length(thres)]

plot(density(thres), main=paste("Distribution of prediction thresholds at ",recall.thres," recall\nThe median values=",median(thres),sep=""), xlab="Prediction Threshold")

# build predictor using a logistic regression
tmp <- data.frame("MAP3K7 HScore"=as.numeric(master.table$HScoreMAP3K7), "CHD1 HScore"=as.numeric(master.table$HScoreCHD1), BCR=factor(hscore.combo$bcr))
model <- glm(BCR ~ ., data = tmp, family=binomial(link='logit'))
class(model) <- c("lr", class(model))
predict.lr <- function(object, newdata, ...)
  predict.glm(object, newdata, type = "response") > median(thres)

tmp.pr <- predict.lr(model,tmp)
tmp.pr.F <- tmp$BCR[tmp.pr==F]
tmp.pr.T <- tmp$BCR[tmp.pr==T]
tp <- length(tmp.pr.F[tmp.pr.F=="BCR"])
fp <- length(tmp.pr.F[tmp.pr.F=="NOBCR"])
fn <- length(tmp.pr.T[tmp.pr.T=="BCR"])
tn <- length(tmp.pr.T[tmp.pr.T=="NOBCR"])
prec <- round(tp / (tp+fp),3)
recall <- round(tp /(tp+fn),3)
acc <- round((tp+tn)/(tp+tn+fp+fn),3)

decisionplot(model, tmp, class = "BCR", main = paste("Logistic Regression, Acc=",acc, ", Prec=",prec, " @ ",recall.thres, " Recall", sep=""), resolution=1000, showgrid=F)

######################## 
# sample the data to determine the threshold to set the prediction limit 0.6 recall
thres <- array()
pats.uni <- unique(hscores$ID)
recall.thres = 0.6
for (q in 1:100){
  df <- data.frame(HScoreMAP3K7=integer(), HScoreCHD1=integer())
  for(i in 1:length(pats.uni)) {
    tmpmat <- hscores[hscores$ID == pats.uni[i],]
    j <- sample(nrow(tmpmat),1)
    df <- rbind(df, data.frame(HScoreMAP3K7=tmpmat$HScoreMAP3K7[j], HScoreCHD1=tmpmat$HScoreCHD1[j]))
    row.names(df)[nrow(df)] <- as.character(tmpmat$ID[j])
  }
  # remove any samples with missing Hscores for ERG, CHD1 or MAP3K7
  df <- df[!is.na(df$HScoreCHD1),]
  df <- df[!is.na(df$HScoreMAP3K7),]
  pats <- intersect(row.names(hscore.combo), row.names(df))
  df <- df[pats,]
  tmp.bcr <- hscore.combo[pats,]
  df <- cbind(df, BCR=tmp.bcr$bcr)
  model <- glm(BCR ~ ., data = df, family=binomial(link='logit'))
  class(model) <- c("lr", class(model))

  model.eval <- rocplotvalues(model, df)
  min.recall <- min(model.eval$recall[model.eval$recall > recall.thres])
  k <- match(min.recall, model.eval$recall)
  thres <- c(thres,seq(0,1,0.01)[k])
}
thres <- thres[2:length(thres)]

plot(density(thres), main=paste("Distribution of prediction thresholds at ",recall.thres," recall\nThe median values=",median(thres),sep=""), xlab="Prediction Threshold")

# build predictor using a logistic regression
tmp <- data.frame("MAP3K7 HScore"=as.numeric(master.table$HScoreMAP3K7), "CHD1 HScore"=as.numeric(master.table$HScoreCHD1), BCR=factor(hscore.combo$bcr))
model <- glm(BCR ~ ., data = tmp, family=binomial(link='logit'))
class(model) <- c("lr", class(model))
predict.lr <- function(object, newdata, ...)
  predict.glm(object, newdata, type = "response") > median(thres)

tmp.pr <- predict.lr(model,tmp)
tmp.pr.F <- tmp$BCR[tmp.pr==F]
tmp.pr.T <- tmp$BCR[tmp.pr==T]
tp <- length(tmp.pr.F[tmp.pr.F=="BCR"])
fp <- length(tmp.pr.F[tmp.pr.F=="NOBCR"])
fn <- length(tmp.pr.T[tmp.pr.T=="BCR"])
tn <- length(tmp.pr.T[tmp.pr.T=="NOBCR"])
prec <- round(tp / (tp+fp),3)
recall <- round(tp /(tp+fn),3)
acc <- round((tp+tn)/(tp+tn+fp+fn),3)

decisionplot(model, tmp, class = "BCR", main = paste("Logistic Regression, Acc=",acc, ", Prec=",prec, " @ ",recall.thres, " Recall", sep=""), resolution=1000, showgrid=F)


######################## 
# sample the data to determine the threshold to set the prediction limit 0.75 recall
thres <- array()
pats.uni <- unique(hscores$ID)
recall.thres = 0.75
for (q in 1:100){
  df <- data.frame(HScoreMAP3K7=integer(), HScoreCHD1=integer())
  for(i in 1:length(pats.uni)) {
    tmpmat <- hscores[hscores$ID == pats.uni[i],]
    j <- sample(nrow(tmpmat),1)
    df <- rbind(df, data.frame(HScoreMAP3K7=tmpmat$HScoreMAP3K7[j], HScoreCHD1=tmpmat$HScoreCHD1[j]))
    row.names(df)[nrow(df)] <- as.character(tmpmat$ID[j])
  }
  # remove any samples with missing Hscores for ERG, CHD1 or MAP3K7
  df <- df[!is.na(df$HScoreCHD1),]
  df <- df[!is.na(df$HScoreMAP3K7),]
  pats <- intersect(row.names(hscore.combo), row.names(df))
  df <- df[pats,]
  tmp.bcr <- hscore.combo[pats,]
  df <- cbind(df, BCR=tmp.bcr$bcr)
  model <- glm(BCR ~ ., data = df, family=binomial(link='logit'))
  class(model) <- c("lr", class(model))

  model.eval <- rocplotvalues(model, df)
  min.recall <- min(model.eval$recall[model.eval$recall > recall.thres])
  k <- match(min.recall, model.eval$recall)
  thres <- c(thres,seq(0,1,0.01)[k])
}
thres <- thres[2:length(thres)]

plot(density(thres), main=paste("Distribution of prediction thresholds at ",recall.thres," recall\nThe median values=",median(thres),sep=""), xlab="Prediction Threshold")

# build predictor using a logistic regression
tmp <- data.frame("MAP3K7 HScore"=as.numeric(master.table$HScoreMAP3K7), "CHD1 HScore"=as.numeric(master.table$HScoreCHD1), BCR=factor(hscore.combo$bcr))
model <- glm(BCR ~ ., data = tmp, family=binomial(link='logit'))
class(model) <- c("lr", class(model))
predict.lr <- function(object, newdata, ...)
  predict.glm(object, newdata, type = "response") > median(thres)

tmp.pr <- predict.lr(model,tmp)
tmp.pr.F <- tmp$BCR[tmp.pr==F]
tmp.pr.T <- tmp$BCR[tmp.pr==T]
tp <- length(tmp.pr.F[tmp.pr.F=="BCR"])
fp <- length(tmp.pr.F[tmp.pr.F=="NOBCR"])
fn <- length(tmp.pr.T[tmp.pr.T=="BCR"])
tn <- length(tmp.pr.T[tmp.pr.T=="NOBCR"])
prec <- round(tp / (tp+fp),3)
recall <- round(tp /(tp+fn),3)
acc <- round((tp+tn)/(tp+tn+fp+fn),3)

decisionplot(model, tmp, class = "BCR", main = paste("Logistic Regression, Acc=",acc, ", Prec=",prec, " @ ",recall.thres, " Recall", sep=""), resolution=1000, showgrid=F)


```


```{r}
sessionInfo()
```